Accurate plant pathogen effector protein classification _ab initio_ with **deepredeff**, an ensemble of convolutional neural networks.

### 2021-03-24 main.Rmd
edits for BMC submission

---
title: "Accurate plant pathogen effector protein classification _ab initio_ with **deepredeff**, an ensemble of convolutional neural networks."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: svm_latex.tex
    toc: no
    includes:
      in_header: preamble.tex
fontsize: 12pt
fontfamily: sans
papersize: a4paper
author:
  - name: Ruth Kristianingsih, Dan MacLean
  - affiliation: The Sainsbury Laboratory, University of East Anglia, Norwich Research Park, Norwich, UK, NR4 7JJ, dan.maclean@tsl.ac.uk
abstract: "Plant pathogens cause billions of dollars of crop loss every year and are a major threat to global food security. Effector proteins are the tools such pathogens use to infect the cell, predicting effectors de novo from sequence is difficult because of the heterogeneity of the sequences. We hypothesised that deep learning classifiers based on Convolutional Neural Networks would be able to identify effectors and deliver new insights. We built a training set of manually curated effector sequences from PHI-Base and used these to train a range of model architectures for classifying bacteria, fungal and oomycete sequences. The best performing classifiers had accuracies from 93 % to 84 %. The models were tested against popular effector detection software on our own test data and data provided with those models. We observed better performance from our models. Specifically our models showed greater accuracy and lower tendencies to call false positives on a secreted protein negative test set and a greater generalisability. We used GRAD-CAM activation map analysis to identify the sequences that activated our CNN-LSTM models and found short but distinct N-terminal regions in each taxon that was indicative of effector sequences. No motifs could be observed in these regions but an analysis of amino acid types indicated differing patterns of enrichment and depletion that varied between taxa. We have produced an R package that will allow others to make easy effector predictions using our models."
geometry: margin=1in
biblio-style: apsr
bibliography: ["r-references.bib"]
graphics: yes
header-includes:
  - \usepackage{amsmath}
  - \usepackage{siunitx}
  - \usepackage{float}
---

@@

# ggplot2 global theme
theme_set(theme_bw())
```

```{r analysis-preferences, echo=FALSE}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(
  cache.extra = knitr::rand_seed,
  message = FALSE,
  echo = FALSE
)
```

# Introduction

@@

# Methods

@@

Sequence data were collected from the PHI-Base database version 4.8 [@phibase] by accessing a text dump of the data prepared on request by the PHI-Base team, the file can be accessed at [https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv). The pipeline in Figure \@ref(fig:effectorcollection) outlines the steps used. We filtered plant effector proteins and their taxonomic groups and collected sequences from UniProt Release 2019_05, using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd). We created a correspondingly sized data set of non-effectors with secretion signals originating in species matched to those from which the effectors were drawn. We downloaded sequences for randomly selected proteins matching these criteria from Ensembl databases [@yates2020] : specifically Ensembl Fungi, Protists and Bacteria manually using the BioMart tools [@smedleybiomart2009]. Since the BioMart tool is not available on Ensembl Bacteria, we downloaded whole proteome protein sequnces from species matched to those from which the effector came using FTP. With these we used SignalP 3.0 [@dyrlovbendtsen2004] in order to filter the secreted sequences and selected accordingly. We used default paramaters from SignalP, except the type of organism group which is `euk` for both fungi and oomycete sequences, and `gram-` or `gram+` for bacteria sequences. Redundant sequences were filtered using BLASTp [@camachoblast2009]. We achieved these steps using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd).

@@

@@

The performance of the trained classifiers is dependent on the quality of the input training data, so it was important that we collected as high a quality set of annotated effectors as possible. To this end we used PHI-Base [@phibase] as our primary sequence origin. Sequences in PHI-Base are human curated from the literature and have therefore been noted in experimental studies. They do not derive from large scale annotations or contain hypothetical or predicted proteins. This attribute makes it ideal for our purposes as the effectors in PHI-Base are those that have been specifically reported as such in the published literature and are not of the class of sequences that are merely suspected of being effectors on the basis of carrying a secretion signal. To collect effector sequences we parsed a whole database text dump of version 4.8 [https://github.com/PHI-base/data](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv), all proteins marked as plant pathogen effectors were filtered and we used the IDs and UniProt IDs to collect the protein sequences from PHI-Base or UniProt if PHI-Base stored only the ID. The sequences and IDs retrieved can be seen in the data file in this manuscript's repository [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/data/getting-data-new/binary-class-data/effector_data.csv](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/data/getting-data-new/binary-class-data/effector_data.csv). Effector sequences were then divided into taxonomic groups as bacterial, fungal or oomycete derived accordingly. In total 190 bacterial effectors from 13 species were collected, 97 fungal effectors from 16 species were collected and 85 oomycete effectors from 6 species were collected (Table \@ref(tab:effsummary) ). The species and effector count in each group can be seen in Tables \@ref(tab:bacefftable), \@ref(tab:fungiefftable) and \@ref(tab:oomefftable).

Sequences for non-effector, secreted proteins were collected using a similar pipeline. Randomly selected proteins from each species carrying secretion signals were extracted from Ensembl databases using the BioMart tool. For each species noted in Tables \@ref(tab:bacefftable), \@ref(tab:fungiefftable) and \@ref(tab:oomefftable) we collected from either the same strain or species an identical number of non-effector, secreted proteins to that in the effector set. This gave us a balanced data set of effector proteins as positive learning examples and non-effector secreted proteins as negative learning examples. Figure \@ref(fig:noneffectorcollection) summarises the process of building the non-effector set, and the full set of sequences and IDs retrieved can be seen in the following data file [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/secreted_data](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/secreted_data).

```{r effectorcollection, fig.width=10, fig.height=10, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt"}
knitr::include_graphics(
  "figures/flowchart_effectorsequence_collection.pdf"
)
```

```{r noneffectorcollection, fig.width=10, fig.height=10, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists"}
knitr::include_graphics(
  "figures/flowchart_secreted_noneffectorsequence_collection.pdf"
)
```

```{r, effsummary}
apa_table(
  readr::read_tsv("tables/effectors_by_path_group.tab"),
  digits = 0,
  align = 'c',
  caption = 'Count of effectors listed in publications curated by PHI-Base used in this study in three major plant pathogen groups'
)
```

## Model Selection and Training
In order to identify a useful classifier we took a randomised hyperparamter search over some likely base model architectures. We selected four base architectures on which to build models for learning. Two of these contained Convolutional Neural Network (CNN) layers followed by either a Long Short Term Memory Layer (LSTM) or a Gated Recurrent Unit (GRU), two contained an Embedding Layer followed by the LSTM or GRU. All models had fully-connected dense layers after this. See Figure \@ref(fig:modelarch).

@@

```{r modelarch, fig.width=10, fig.height=10, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C: LSTM-Embedding model; D: GRU-Embedding model"}
knitr::include_graphics("figures/model_arch.pdf")

@@

```{r, acclossfinal}
 readr::read_tsv("tables/acc_loss_final.tab") %>%
  apa_table(            digits = 3,
  align = 'c',
  caption = 'Accuracy and loss values from best performing parameters values for each model'
         )
```

@@

Final training progressions for each model in each taxon can be seen in Figure \@ref(fig:trainingprogress). We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4  to 97.4, for the four fungal models we observed accuracy in the range  60.5 to 84.2 and for the four oomycete models we observed accuracy from 64.7 to 82.3, reported in Fig \@ref(fig:overallmodel). All the models we generated had high statistical power and can accurately and reliably classify effectors from other secreted proteins in that taxon.

@@

```{r trainingprogress, fig.height= 10, fig.width= 10, fig.cap="Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set"}
# knitr::include_graphics("figures/final_train_progress.png")

@@

```{r overallmodel, fig.cap="Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences."}
# apa_table(
#   readr::read_tsv("tables/final_model_performance.tab"),
#   digits = 3,
#   align = 'c',
#   caption = 'Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.'
# )
readr::read_tsv("tables/final_model_performance_tidy.tab") %>%
  ggplot2::ggplot() +
  ggplot2::aes(
    x = Model,
    y = Accuracy,
    color = Accuracy
  ) +
  ggplot2::geom_point() +
  ggplot2::scale_color_viridis_c(begin = 0.65, end = 0.1) +
  ggplot2::facet_wrap(~ Taxon, scales = "free_x") +
  ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

Classification correlations between the different model architectures were high and positive in the bacterial sequence trained model's calls, in the range 0.8 to 0.88, see Figure \@ref(fig:confusionmodelcorrs). CNN-GRU and LSTM-Embedding showed identical prediction sets. We observed similar levels of correlation  in the CNN-LSTM, GRU-Embedding and CNN-GRU fungal sequence trained model, in the range 0.79 to 0.9 ; though there was a significantly lower range of correlations with the LSTM-Embedding which were in the range 0.36 to 0.51. The models trained on oomycete sequences all showed this lower range of correlations, in the range 0.30 to 0.65. The higher correlation across bacterial trained models is likely from a mixture of the larger training set size and a greater uniformity of the sequences themselves. For the fungal sequence trained models we can see that the LSTM-Embedding model does not perform as well as the others. The oomycete sequence trained models all show a lower range correlation reflecting the likely less uniform and smaller training set. It is clear that, particularly for the fungal and oomycete models, each architecture is capturing separate aspects of the sequences and classifying on those with slightly varying levels of success.

```{r confusionmodelcorrs, fig.height = 8, fig.width = 8, fig.cap = "Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data; and Pearson correlations between classifications made by different models. A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences.", message = FALSE, echo=FALSE}

@@

We examined the usefulness of combining the predictions of the different model architectures using an ensemble function that takes the vectors of classifications of each model architecture as input. We performed the classification of the hold-out test data set using the ensembled models and the results can be seen in Figure \@ref(fig:ensembleresults). With the models trained in bacterial sequences we observed an increase in classification accuracy over the best model, up to 0.99 for both ensemble functions. However, with the fungal and oomycete models we observed decreases relative to the best single model in both cases due to the higher accuracy of the CNN-LSTM model being diluted by the combined inacuracy of the other model architectures. Examining the overlaps in classifications between the CNN-LSTM and CNN-GRU/LSTM-Embedding respectively showed that the two lesser performing models were not simply predicting subsets of the CNN-LSTM model, in both cases the lesser models were able to identify three effectors correctly that were missed by the generally stronger models. This indicates that the weaker models may be classifying on some patterns missed by the CNN-LSTM model.

```{r ensembleresults, fig.width=8, fig.height=4, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences."}
# apa_table(
#   readr::read_tsv("tables/ensemble_model_performance.tab"),
#   digits = 3,
#   align = 'c',
#   caption = 'Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.'
readr::read_tsv("tables/ensemble_model_performance_tidy.tab") %>%
  ggplot2::ggplot() +
  ggplot2::aes(x = Combination, y = Accuracy) +
  ggplot2::geom_point(ggplot2::aes(colour = EnsembleFunction,
                                   shape = EnsembleFunction), size = 2) +
  ggplot2::scale_shape_manual(values = c("Ensemble Voting" = 19, "Ensemble Weighted" = 17)) +
  ggplot2::scale_color_viridis_d(begin = 0.65, end = 0.1) +
  ggplot2::labs(color = "Ensemble Function", shape = "Ensemble Function") +
  ggplot2::facet_wrap(~ Taxon, scales = "free_x") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

@@

Given the accuracies of the above we selected the ensemble bacterial model and the CNN-LSTM fungal and oomycete models to evaluate the performance of our models against widely used effector identification software. We compared against predictions from the bacterial effector prediction programs DeepT3 [@deept3] and EffectiveT3 classification module for plant-associated bacteria 1.0.1 [@effectivet3], the fungal effector prediction programs EffectorP 1.0 and 2.0 [@effectorp, @effectorp2] and the oomycete effector predictor EffectR [@effectR]. Each comparison was carried out using the respective hold-out test sequence set for each taxon. For all taxa we observed greater Accuracy and F1 scores from our models than the established software, as shown in Figure \@ref(fig:vsothersmetrics). This was particularly marked in the F1 score, which incorporates a measure of the incorrect calls. Absolute improvements were up to 15 %  in bacterial sequences, 15 % in fungal sequences and 20 %  in the oomycete sequences. The confusion matrices in Figure \@ref(fig:vsotherscfmat) show that accuracy and F1 score was compromised in all the established tools by the tendency of them all to misclassify true effectors as not effectors.  All the established software classifiers we tested show lower sensitivity than the models we have developed here.

We also evaluated the deep learning models we have developed on the training and hold-out validation sequences used to train the previous methods. We calculated the proportion of the effectors in the training set that the tools could find on their respective training and validation sets, according to availability. The bacterial tools EffectiveT3 and DeepT3 showed lower proportion found than our Ensemble Weighted model, as in Figure \@ref(fig:owndata)A, consistent with the observation that our Ensemble Weighted model performed more strongly on the validation set that we generated. Interestingly, both versions of EffectorP found a greater proportion of the effectors in the EffectorP provided training sets than our CNN-LSTM model, but in the unseen validation data provided with EffectorP 2, all three models performed identically (Figure \@ref(fig:owndata)B ). The EffectorP 1 and 2 scores on validation data are well below the scores for the training data, a result that is usually interpreted as being evidence of an overfitted and less generalisable model. Our CNN-LSTM model for fungi showed similar scores across the training and validation set indicating a greater generability and equivalent power. Only six effector sequences are provided as examples with the oomycete specific effector finder effectR. As this is not a trainable model in the same sense as the others, no large training set is needed. We attempted classification with these and effectR was able to classify 5 of the 6, whereas our CNN-LSTM model for oomycetes classified 3 of the 6 ( Figure \@ref(fig:owndata)C ).

@@

```{r, owndata, fig.height = 8, fig.cap="Performance of models on training and validation datasets used to train other effector prediction tools. A: Tools trained on bacterial sequences; B: Tools trained on fungal sequences; C: Tools trained on oomycete sequences."}
d <- readr::read_csv(here::here("data","processed", "owndatasets.csv"))

@@

To make the models useful for developers in their own analytic pipelines we have provided an R package that provides a useful interface to the models. The package and installation instructions are available from GitHub [https://ruthkr.github.io/deepredeff](https://ruthkr.github.io/deepredeff)

#########################################################
## Start code fragments
##
##
#########################################################

@@

>>>>

---
title: "Accurate plant pathogen effector protein classification _ab initio_ with **deepredeff**, an ensemble of convolutional neural networks."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: svm_latex.tex
    toc: no
    includes:
      in_header: preamble.tex
fontsize: 12pt
fontfamily: sans
papersize: a4paper
author:
  - name: Ruth Kristianingsih, Dan MacLean
  - affiliation: The Sainsbury Laboratory, University of East Anglia, Norwich Research Park, Norwich, UK, NR4 7JJ, dan.maclean@tsl.ac.uk
abstract: ""
geometry: margin=1in
biblio-style: apsr
bibliography: ["r-references.bib"]
graphics: yes
header-includes:
  - \usepackage{amsmath}
  - \usepackage{siunitx}
  - \usepackage{float}
---

@@

# ggplot2 global theme
theme_set(theme_bw())

fig_caption <- captioner::captioner(prefix = "Figure ")
fig_caption("effectorcollection", "Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt")

fig_caption("noneffectorcollection", "Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists")

fig_caption("modelarch", "Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C: LSTM-Embedding model; D: GRU-Embedding model")

fig_caption("trainingprogress", "Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set")

fig_caption("overallmodel", "Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.")

fig_caption("confusionmodelcorrs", "Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data; and Pearson correlations between classifications made by different models. A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences.")

fig_caption("ensembleresults", "Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.")

fig_caption("vsothersmetrics", "Accuracy and F1 score of the classifications made by our models compared with those made by widely used software")

fig_caption("vsotherscfmat", "Confusion matrices of classifications made by our models compared with those made by widely used software. A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences.")

fig_caption("owndata", "Performance of models on training and validation datasets used to train other effector prediction tools. A: Tools trained on bacterial sequences; B: Tools trained on fungal sequences; C: Tools trained on oomycete sequences.")

fig_caption("activations", "Grad-CAM activation profiles of CNN-LSTM models  when run with all datasets; and sequence logo plots of sequence in the activation region within the dotted lines for effector data (left panel) and non-effector data (right panel). A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences.")

tbl_caption <- captioner::captioner(prefix = "Table ")
tbl_caption("effsummary", 'Count of effectors listed in publications curated by PHI-Base used in this study in three major plant pathogen groups')

tbl_caption("aclossfinal", 'Accuracy and loss values from best performing parameters values for each model')

tbl_caption("phyperaatype", 'Enrichment analysis of amino acid types. Proportions of each amino acid type in the activation region of All Sequences : Effector Sequences, with (Probability). Probability by hypergeometric test of the observed number of amino acid type in the region of activation in effector sequences in our compiled sequence set relative to a background of the same region in all sequences in that taxon.')
```

```{r analysis-preferences, echo=FALSE}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(
  cache.extra = knitr::rand_seed,
  message = FALSE,
  echo = FALSE
)
```
\newpage
# Abstract

## Background:

Plant pathogens cause billions of dollars of crop loss every year and are a major threat to global food security. Effector proteins are the tools such pathogens use to infect the cell, predicting effectors de novo from sequence is difficult because of the heterogeneity of the sequences. We hypothesised that deep learning classifiers based on Convolutional Neural Networks would be able to identify effectors and deliver new insights.

## Results:

We created a training set of manually curated effector sequences from PHI-Base and used these to train a range of model architectures for classifying bacteria, fungal and oomycete sequences. The best performing classifiers had accuracies from 93 % to 84 %. The models were tested against popular effector detection software on our own test data and data provided with those models. We observed better performance from our models. Specifically our models showed greater accuracy and lower tendencies to call false positives on a secreted protein negative test set and a greater generalisability. We used GRAD-CAM activation map analysis to identify the sequences that activated our CNN-LSTM models and found short but distinct N-terminal regions in each taxon that was indicative of effector sequences. No motifs could be observed in these regions but an analysis of amino acid types indicated differing patterns of enrichment and depletion that varied between taxa.

## Conclusions:

Small training sets can be used effectively to train highly accurate and sensitive deep learning models without need for the operator to know anything other than sequence and without arbitrary decisions made about what sequence features or physico-chemical properties are important. Biological insight on subsequences important for classification can be achieved by examining the activations in the model

\newpage
# Background

@@

# Implementation

@@

Sequence data were collected from the PHI-Base database version 4.8 [@phibase] by accessing a text dump of the data prepared on request by the PHI-Base team, the file can be accessed at [https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv). The pipeline in Figure \@ref(effectorcollection) outlines the steps used. We filtered plant effector proteins and their taxonomic groups and collected sequences from UniProt Release 2019_05, using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd). We created a correspondingly sized data set of non-effectors with secretion signals originating in species matched to those from which the effectors were drawn. We downloaded sequences for randomly selected proteins matching these criteria from Ensembl databases [@yates2020] : specifically Ensembl Fungi, Protists and Bacteria manually using the BioMart tools [@smedleybiomart2009]. Since the BioMart tool is not available on Ensembl Bacteria, we downloaded whole proteome protein sequnces from species matched to those from which the effector came using FTP. With these we used SignalP 3.0 [@dyrlovbendtsen2004] in order to filter the secreted sequences and selected accordingly. We used default paramaters from SignalP, except the type of organism group which is `euk` for both fungi and oomycete sequences, and `gram-` or `gram+` for bacteria sequences. Redundant sequences were filtered using BLASTp [@camachoblast2009]. We achieved these steps using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd).

@@

To make our models useful for developers in their own analytic pipelines we have provided an R package that provides a useful interface to the models. The package and installation instructions are available from GitHub [https://ruthkr.github.io/deepredeff](https://ruthkr.github.io/deepredeff) and CRAN [https://cran.r-project.org/package=deepredeff](https://cran.r-project.org/package=deepredeff)

@@

The performance of the trained classifiers is dependent on the quality of the input training data, so it was important that we collected as high a quality set of annotated effectors as possible. To this end we used PHI-Base [@phibase] as our primary sequence origin. Sequences in PHI-Base are human curated from the literature and have therefore been noted in experimental studies. They do not derive from large scale annotations or contain hypothetical or predicted proteins. This attribute makes it ideal for our purposes as the effectors in PHI-Base are those that have been specifically reported as such in the published literature and are not of the class of sequences that are merely suspected of being effectors on the basis of carrying a secretion signal. To collect effector sequences we parsed a whole database text dump of version 4.8 [https://github.com/PHI-base/data](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv), all proteins marked as plant pathogen effectors were filtered and we used the IDs and UniProt IDs to collect the protein sequences from PHI-Base or UniProt if PHI-Base stored only the ID (see `r fig_caption("effectorcollection", display="cite")`. The sequences and IDs retrieved can be seen in the data file in this manuscript's repository [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/data/getting-data-new/binary-class-data/effector_data.csv](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/data/getting-data-new/binary-class-data/effector_data.csv). Effector sequences were then divided into taxonomic groups as bacterial, fungal or oomycete derived accordingly. In total 190 bacterial effectors from 13 species were collected, 97 fungal effectors from 16 species were collected and 85 oomycete effectors from 6 species were collected (Table \@ref(tab:effsummary) ). The species and effector count in each group can be seen in Tables \@ref(tab:bacefftable), \@ref(tab:fungiefftable) and \@ref(tab:oomefftable).

Sequences for non-effector, secreted proteins were collected using a similar pipeline. Randomly selected proteins from each species carrying secretion signals were extracted from Ensembl databases using the BioMart tool. For each species noted in Tables \@ref(tab:bacefftable), \@ref(tab:fungiefftable) and \@ref(tab:oomefftable) we collected from either the same strain or species an identical number of non-effector, secreted proteins to that in the effector set. This gave us a balanced data set of effector proteins as positive learning examples and non-effector secreted proteins as negative learning examples. `r fig_caption("noneffectorcollection", display="cite")` summarises the process of building the non-effector set, and the full set of sequences and IDs retrieved can be seen in the following data file [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/secreted_data](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/secreted_data).

```{r effectorcollection, fig.width=10, fig.height=10, out.width='50%', fig.align='center', fig.cap=fig_caption("effectorcollection",display="cite")}
knitr::include_graphics(
  "figures/flowchart_effectorsequence_collection.pdf"
)
```

```{r noneffectorcollection, fig.width=10, fig.height=10, fig.cap=fig_caption("noneffectorcollection",display="cite")}
knitr::include_graphics(
  "figures/flowchart_secreted_noneffectorsequence_collection.pdf"
)
```

```{r, effsummary}
apa_table(
  readr::read_tsv("tables/effectors_by_path_group.tab"),
  digits = 0,
  align = 'c',
  caption = " "
)
```

## Model Selection and Training
In order to identify a useful classifier we took a randomised hyperparamter search over some likely base model architectures. We selected four base architectures on which to build models for learning. Two of these contained Convolutional Neural Network (CNN) layers followed by either a Long Short Term Memory Layer (LSTM) or a Gated Recurrent Unit (GRU), two contained an Embedding Layer followed by the LSTM or GRU. All models had fully-connected dense layers after this. See  `r fig_caption("modelarch", display="cite")`.

@@

```{r modelarch, fig.width=10, fig.height=10, fig.cap=fig_caption("modelarch",display="cite")}
knitr::include_graphics("figures/model_arch.pdf")

@@

```{r, acclossfinal}
 readr::read_tsv("tables/acc_loss_final.tab") %>%
  apa_table(            digits = 3,
  align = 'c',
  caption = ' '
         )
```

@@

Final training progressions for each model in each taxon can be seen in `r fig_caption("trainingprogress", display="cite")`. We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4  to 97.4, for the four fungal models we observed accuracy in the range  60.5 to 84.2 and for the four oomycete models we observed accuracy from 64.7 to 82.3, reported in `r fig_caption("overallmodel", display="cite")`. All the models we generated had high statistical power and can accurately and reliably classify effectors from other secreted proteins in that taxon.

@@

```{r trainingprogress, fig.height= 10, fig.width= 10, fig.cap=fig_caption("trainingprogress",display="cite")}
# knitr::include_graphics("figures/final_train_progress.png")

@@

```{r overallmodel, fig.cap=fig_caption("overallmodel",display="cite")}
# apa_table(
#   readr::read_tsv("tables/final_model_performance.tab"),
#   digits = 3,
#   align = 'c',
#   caption = 'Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.'
# )
readr::read_tsv("tables/final_model_performance_tidy.tab") %>%
  ggplot2::ggplot() +
  ggplot2::aes(
    x = Model,
    y = Accuracy,
    color = Accuracy
  ) +
  ggplot2::geom_point() +
  ggplot2::scale_color_viridis_c(begin = 0.65, end = 0.1) +
  ggplot2::facet_wrap(~ Taxon, scales = "free_x") +
  ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

Classification correlations between the different model architectures were high and positive in the bacterial sequence trained model's calls, in the range 0.8 to 0.88, see  `r fig_caption("confusionmodelcorrs", display="cite")`. CNN-GRU and LSTM-Embedding showed identical prediction sets. We observed similar levels of correlation  in the CNN-LSTM, GRU-Embedding and CNN-GRU fungal sequence trained model, in the range 0.79 to 0.9 ; though there was a significantly lower range of correlations with the LSTM-Embedding which were in the range 0.36 to 0.51. The models trained on oomycete sequences all showed this lower range of correlations, in the range 0.30 to 0.65. The higher correlation across bacterial trained models is likely from a mixture of the larger training set size and a greater uniformity of the sequences themselves. For the fungal sequence trained models we can see that the LSTM-Embedding model does not perform as well as the others. The oomycete sequence trained models all show a lower range correlation reflecting the likely less uniform and smaller training set. It is clear that, particularly for the fungal and oomycete models, each architecture is capturing separate aspects of the sequences and classifying on those with slightly varying levels of success.

```{r confusionmodelcorrs, fig.height = 8, fig.width = 8, fig.cap = fig_caption("confusionmodelcorrs",display="cite"), message = FALSE, echo=FALSE}

@@

We examined the usefulness of combining the predictions of the different model architectures using an ensemble function that takes the vectors of classifications of each model architecture as input. We performed the classification of the hold-out test data set using the ensembled models and the results can be seen in `r fig_caption("ensembleresults", display="cite")`. With the models trained in bacterial sequences we observed an increase in classification accuracy over the best model, up to 0.99 for both ensemble functions. However, with the fungal and oomycete models we observed decreases relative to the best single model in both cases due to the higher accuracy of the CNN-LSTM model being diluted by the combined inacuracy of the other model architectures. Examining the overlaps in classifications between the CNN-LSTM and CNN-GRU/LSTM-Embedding respectively showed that the two lesser performing models were not simply predicting subsets of the CNN-LSTM model, in both cases the lesser models were able to identify three effectors correctly that were missed by the generally stronger models. This indicates that the weaker models may be classifying on some patterns missed by the CNN-LSTM model.

```{r ensembleresults, fig.width=8, fig.height=4, fig.cap=fig_caption("ensembleresults",display="cite")}
# apa_table(
#   readr::read_tsv("tables/ensemble_model_performance.tab"),
#   digits = 3,
#   align = 'c',
#   caption = 'Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.'
readr::read_tsv("tables/ensemble_model_performance_tidy.tab") %>%
  ggplot2::ggplot() +
  ggplot2::aes(x = Combination, y = Accuracy) +
  ggplot2::geom_point(ggplot2::aes(colour = EnsembleFunction,
                                   shape = EnsembleFunction), size = 2) +
  ggplot2::scale_shape_manual(values = c("Ensemble Voting" = 19, "Ensemble Weighted" = 17)) +
  ggplot2::scale_color_viridis_d(begin = 0.65, end = 0.1) +
  ggplot2::labs(color = "Ensemble Function", shape = "Ensemble Function") +
  ggplot2::facet_wrap(~ Taxon, scales = "free_x") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

@@

Given the accuracies of the above we selected the ensemble bacterial model and the CNN-LSTM fungal and oomycete models to evaluate the performance of our models against widely used effector identification software. We compared against predictions from the bacterial effector prediction programs DeepT3 [@deept3] and EffectiveT3 classification module for plant-associated bacteria 1.0.1 [@effectivet3], the fungal effector prediction programs EffectorP 1.0 and 2.0 [@effectorp, @effectorp2] and the oomycete effector predictor EffectR [@effectR]. Each comparison was carried out using the respective hold-out test sequence set for each taxon. For all taxa we observed greater Accuracy and F1 scores from our models than the established software, as shown in `r fig_caption("vsothersmetrics", display="cite")`. This was particularly marked in the F1 score, which incorporates a measure of the incorrect calls. Absolute improvements were up to 15 %  in bacterial sequences, 15 % in fungal sequences and 20 %  in the oomycete sequences. The confusion matrices in  `r fig_caption("vsotherscfmat", display="cite")` show that accuracy and F1 score was compromised in all the established tools by the tendency of them all to misclassify true effectors as not effectors.  All the established software classifiers we tested show lower sensitivity than the models we have developed here.

We also evaluated the deep learning models we have developed on the training and hold-out validation sequences used to train the previous methods. We calculated the proportion of the effectors in the training set that the tools could find on their respective training and validation sets, according to availability. The bacterial tools EffectiveT3 and DeepT3 showed lower proportion found than our Ensemble Weighted model, as in `r fig_caption("owndata", display="cite")`A, consistent with the observation that our Ensemble Weighted model performed more strongly on the validation set that we generated. Interestingly, both versions of EffectorP found a greater proportion of the effectors in the EffectorP provided training sets than our CNN-LSTM model, but in the unseen validation data provided with EffectorP 2, all three models performed identically ( `r fig_caption("owndata", display="cite")`B ). The EffectorP 1 and 2 scores on validation data are well below the scores for the training data, a result that is usually interpreted as being evidence of an overfitted and less generalisable model. Our CNN-LSTM model for fungi showed similar scores across the training and validation set indicating a greater generability and equivalent power. Only six effector sequences are provided as examples with the oomycete specific effector finder effectR. As this is not a trainable model in the same sense as the others, no large training set is needed. We attempted classification with these and effectR was able to classify 5 of the 6, whereas our CNN-LSTM model for oomycetes classified 3 of the 6 (  `r fig_caption("owndata", display="cite")`C ).

@@

```{r, owndata, fig.height = 8, fig.cap=fig_caption("owndata",display="cite")}
d <- readr::read_csv(here::here("data","processed", "owndatasets.csv"))

@@

@@

# List Of Abbreviations

ML: Machine Learning,
CNN: Convolutional Neural Network,
MPMI: Molecular Plant Microbe Interactions,
LSTM: Long Short Term Memory,
GRU: Gated Recurrent Unit,
CPU: Central Processing Unit,
GPU: Graphics Processing Unit,
TM: Trademark,
TP: True Positive,
FP: False Positive,
FN: False Negative,
TN: True Negative,
Grad-CAM: Gradient-weighted Class Activation Mapping,
ID: Identifier,
FFT: Fast Fourier Transform,

# Declarations

## Availability of data and material

The datasets generated and/or analysed during the current study are available in the 'ruth-effectors-prediction' repository, https://github.com/TeamMacLean/ruth-effectors-prediction. Individual datasets location within this repository are listed per dataset in the Results section

## Competing Interests

We have no competing interests.

## Funding
